Asthma Cohort Descriptive Epidemiology

### 2017-12-08 r_scripts/analysis/asthma_cohort_descriptive_epi.Rmd
added MSA in to descriptive summary.

```{r cohort_import}
# read path
read_path <- "../../data/health/2013-oregon_asthma_cohort.csv"
# read in asthma cohort (reading a million rows for now)
asthma_cohort <- data.table::fread(read_path, 
  # read options
  header = T, #nrows = 10000, # removing nrows read 
  colClasses = rep("character", 72)) %>% 
  # filter to oregon
  filter(STATE == "OR") %>% 
  # replace *NULL* with NA
  mutate_all(funs(replace(., .== "*NULL*" | . == "", NA))) %>% 
  # create visit type indicator 
  mutate(visit_type = as.factor(
    case_when((dx1 %in% asthma_icd9) ~ "dx_asthma_primary",
    # asthma but not as a primary
    !(dx1 %in% asthma_icd9) & (dx2 %in% asthma_icd9 | dx3 %in% asthma_icd9 | 
      dx4 %in% asthma_icd9 | dx5 %in% asthma_icd9 | dx6 %in% asthma_icd9 | 
      dx7 %in% asthma_icd9 | dx8 %in% asthma_icd9 | dx9 %in% asthma_icd9 | 
      dx10 %in% asthma_icd9 | dx11 %in% asthma_icd9 | dx12 %in% asthma_icd9 | 
      dx13 %in% asthma_icd9) ~ "dx_asthma_not_primary",
    # other dx 
    !(dx1 %in% asthma_icd9 | dx2 %in% asthma_icd9 | dx3 %in% asthma_icd9 | 
      dx4 %in% asthma_icd9 | dx5 %in% asthma_icd9 | dx6 %in% asthma_icd9 | 
      dx7 %in% asthma_icd9 | dx8 %in% asthma_icd9 | dx9 %in% asthma_icd9 | 
      dx10 %in% asthma_icd9 | dx11 %in% asthma_icd9 | dx12 %in% asthma_icd9 | 
      dx13 %in% asthma_icd9) & is.na(ndc) ~ "dx_other",
    # ndc code
    (ndc %in% saba_ndc) ~ "pharm_saba",
    !is.na(ndc) & !(ndc %in% saba_ndc) ~ "pharm_other"))) %>% 
  # convert date variables to dates
  mutate_at(vars(contains("date")), funs(as.Date(.,format="%Y-%m-%d"))) %>%
  # adding and converting some variables
  mutate(los = as.numeric(los),
         # create age variable
         age = 2013 - as.numeric(yob),
         # code in race text to numeric race
         race_factor = case_when(race == 1 ~ "White",
                                 race == 2 ~ "Black",
                                 race == 3 ~ "American Indian",
                                 race == 4 ~ "Asian",
                                 race == 5 ~ "Pacific Islander",
                                 race == 6 ~ "Other", 
                                 race == 7 ~ "Two or more",
                                 race == 9 ~ "Unknown"),
         ethnicity = case_when(ethn == "1" ~ "Hispanic",
                               ethn == "2" ~ "Not Hispanic",
                               ethn == "3" ~ "Unknown")) %>% 
  # join in place of service text
  left_join(place_of_service, by = "pos") %>% 
  # simpler place of service grouping
  mutate(pos_simple = as.factor(case_when(pos == "01" ~ "Pharmacy",
    pos == "03" ~ "School",
    pos == "11" ~ "Office",
    pos == "20" ~ "Urgent Care", 
    pos == "21" ~ "Inpatient Hospital",
    pos == "22" ~ "Outpatient Hospital",
    pos == "23" ~ "Emergency Room Hospital",
    pos == "41" | pos == "42" ~ "Ambulance",
    pos == "71" | pos == "72" ~ "Public Health Clinic",
    pos == "49" ~ "Independent Clinic", 
    TRUE ~ "Other")))
```

@@

>>>>

```{r cohort_import}
# read path
read_path <- "../../data/health/2013-oregon_asthma_cohort.csv"
# read in asthma cohort (reading a million rows for now)
asthma_cohort <- data.table::fread(read_path, 
  # read options
  header = T, #nrows = 10000, # removing nrows read 
  colClasses = rep("character", 72)) %>% 
  # filter to oregon
  filter(STATE == "OR") %>% 
  # replace *NULL* with NA
  mutate_all(funs(replace(., .== "*NULL*" | . == "", NA))) %>% 
  # create visit type indicator 
  mutate(visit_type = as.factor(
    case_when((dx1 %in% asthma_icd9) ~ "dx_asthma_primary",
    # asthma but not as a primary
    !(dx1 %in% asthma_icd9) & (dx2 %in% asthma_icd9 | dx3 %in% asthma_icd9 | 
      dx4 %in% asthma_icd9 | dx5 %in% asthma_icd9 | dx6 %in% asthma_icd9 | 
      dx7 %in% asthma_icd9 | dx8 %in% asthma_icd9 | dx9 %in% asthma_icd9 | 
      dx10 %in% asthma_icd9 | dx11 %in% asthma_icd9 | dx12 %in% asthma_icd9 | 
      dx13 %in% asthma_icd9) ~ "dx_asthma_not_primary",
    # other dx 
    !(dx1 %in% asthma_icd9 | dx2 %in% asthma_icd9 | dx3 %in% asthma_icd9 | 
      dx4 %in% asthma_icd9 | dx5 %in% asthma_icd9 | dx6 %in% asthma_icd9 | 
      dx7 %in% asthma_icd9 | dx8 %in% asthma_icd9 | dx9 %in% asthma_icd9 | 
      dx10 %in% asthma_icd9 | dx11 %in% asthma_icd9 | dx12 %in% asthma_icd9 | 
      dx13 %in% asthma_icd9) & is.na(ndc) ~ "dx_other",
    # ndc code
    (ndc %in% saba_ndc) ~ "pharm_saba",
    !is.na(ndc) & !(ndc %in% saba_ndc) ~ "pharm_other")),
    # adding names of metro area
    metroarea = case_when(
      MSA == 13460 ~ "Bend-Redmond",
      MSA == 18700 ~ "Corvallis",
      MSA == 21660 ~ "Eugene-Springfield",
      MSA == 32780 ~ "Medford",
      MSA == 38900 ~ "Portland-Vancouver-Beaverton",
      MSA == 41420 ~ "Salem",
      MSA == 41999 ~ "Not in MSA"),
    los = as.numeric(los),
         # create age variable
         age = 2013 - as.numeric(yob),
         # code in race text to numeric race
         race_factor = case_when(race == 1 ~ "White",
                                 race == 2 ~ "Black",
                                 race == 3 ~ "American Indian",
                                 race == 4 ~ "Asian",
                                 race == 5 ~ "Pacific Islander",
                                 race == 6 ~ "Other", 
                                 race == 7 ~ "Two or more",
                                 race == 9 ~ "Unknown"),
         ethnicity = case_when(ethn == "1" ~ "Hispanic",
                               ethn == "2" ~ "Not Hispanic",
                               ethn == "3" ~ "Unknown")) %>% 
  # convert date variables to dates
  mutate_at(vars(contains("date")), funs(as.Date(.,format="%Y-%m-%d"))) %>%
  # join in place of service text
  left_join(place_of_service, by = "pos") %>% 
  # simpler place of service grouping
  mutate(pos_simple = as.factor(case_when(pos == "01" ~ "Pharmacy",
    pos == "03" ~ "School",
    pos == "11" ~ "Office",
    pos == "20" ~ "Urgent Care", 
    pos == "21" ~ "Inpatient Hospital",
    pos == "22" ~ "Outpatient Hospital",
    pos == "23" ~ "Emergency Room Hospital",
    pos == "41" | pos == "42" ~ "Ambulance",
    pos == "71" | pos == "72" ~ "Public Health Clinic",
    pos == "49" ~ "Independent Clinic", 
    TRUE ~ "Other")))
```

@@

#### Metropolitan Statistical Area

MSA n and percentage.
```{r msa}
xtabs(~metroarea, asthma_demographic)
round(prop.table(xtabs(~metroarea, asthma_demographic)),3)*100
```
